---
title: "Comparing Hybrid Lines with CPM Normalization"
output: html_document
date: '2024-07-09'
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We first compare human and chimp genes that show significant dispersion for two cell types: "Cardiomyocytes" and "Undifferentiated 1". We perform a separate analysis based on the replicate (1, 2, or 3).
```{r,warning=FALSE}
require(stringr)
replicate = 'Rep3'
cell_type = "Undifferentiated1"
#setwd("/project2/gilad/awchen55/differentialDispersion/")
path = '/project2/gilad/awchen55/differentialDispersion/data/hybrid_lines_cpm_normalized/'
#input_name = paste0(path,'chimp_ase_cpm_', replicate)
#output_name_c = paste0(path,'chimp_ase_cpm_', replicate,'_',cell_type)
#output_name_h = paste0(path, 'human_ase_cpm_', replicate, '_', cell_type)
reps <- c("Rep1","Rep2","Rep3")
cell_type <- c()
replicate <- c()
cell_count <- c()

for (rep in reps){
  input_name = paste0(path, 'chimp_ase_cpm_', rep)
  data_h <- read.csv(paste0(input_name,"_cell_type_counts.csv"),row.names="X")
  rownames(data_h) <- data_h$labels
  rownames(data_h)<-str_replace_all(rownames(data_h), c(" " = "."))
  data_h <- as.data.frame(t(data_h))
  data_h = data_h[-1,]
  cell_count <- c(cell_count,max(data_h$Undifferentiated.1_total))
  cell_type <-c(cell_type, "Undifferentiated 1")
  replicate <- c(replicate,rep)
  cell_count <- c(cell_count,max(data_h$Cardiomyocytes_total))
  cell_type <-c(cell_type, "Cardiomyocytes")
  replicate <- c(replicate,rep)
  
}


```

We look at summary of cell counts for each replicate and cell type.
```{r}

cbind(cell_type,replicate,cell_count)
```

We identify the number of significantly dispersed genes for "Undifferentiated 1".
```{r}
cell_type = "Undifferentiated1"
underdispersed <- c()
overdispersed <- c()
for (rep in reps){
  output_name_c = paste0(path,'chimp_ase_cpm_', rep,'_',cell_type)
  output_name_h = paste0(path, 'human_ase_cpm_', rep, '_', cell_type)
  data_c <- read.csv(paste0(output_name_c,"_dispersion.csv"))
  data_h <- read.csv(paste0(output_name_h,"_dispersion.csv"))
  
  underdispersed <- c(underdispersed,nrow(data_c[(data_c$z_score<0) & (data_c$p_value_adj<0.05) & (data_c$cell_count>10) & (data_c$est>0),]))
  overdispersed <- c(overdispersed,nrow(data_c[(data_c$z_score>0) & (data_c$p_value_adj<0.05) & (data_c$cell_count>10) & (data_c$est>0),]))
  
  underdispersed <- c(underdispersed,nrow(data_h[(data_h$z_score<0) & (data_h$p_value_adj<0.05) & (data_h$cell_count>10) & (data_h$est>0),]))
  overdispersed <- c(overdispersed,nrow(data_h[(data_h$z_score>0) & (data_h$p_value_adj<0.05) & (data_h$cell_count>10) & (data_h$est>0),]))
  
}

species <- c("chimp","human","chimp","human","chimp","human")
replicate <- c("Rep1","Rep1","Rep2","Rep2","Rep3","Rep3")


cbind(species,replicate,underdispersed,overdispersed)
```


We identify the number of significantly dispersed genes for "Cardiomyocytes".
```{r}
cell_type = "Cardiomyocytes"
underdispersed <- c()
overdispersed <- c()
overlap_underdispersed <-c()
overlap_overdispersed <-c()
for (rep in reps){
  output_name_c = paste0(path,'chimp_ase_cpm_', rep,'_',cell_type)
  output_name_h = paste0(path, 'human_ase_cpm_', rep, '_', cell_type)
  data_c <- read.csv(paste0(output_name_c,"_dispersion.csv"))
  data_h <- read.csv(paste0(output_name_h,"_dispersion.csv"))
  
  # number of significant genes in chimp
  underdispersed <- c(underdispersed,nrow(data_c[(data_c$z_score<0) & (data_c$p_value_adj<0.05) & (data_c$cell_count>10) & (data_c$est>0),]))
  overdispersed <- c(overdispersed,nrow(data_c[(data_c$z_score>0) & (data_c$p_value_adj<0.05) & (data_c$cell_count>10) & (data_c$est>0),]))
  
  # list of genes in chimp
  under_c_list <- data_c[(data_c$z_score<0) & (data_c$p_value_adj<0.05) & (data_c$cell_count>10) & (data_c$est>0),]$X
  over_c_list <- data_c[(data_c$z_score>0) & (data_c$p_value_adj<0.05) & (data_c$cell_count>10) & (data_c$est>0),]$X
  
  # number of significant genes in human
  underdispersed <- c(underdispersed,nrow(data_h[(data_h$z_score<0) & (data_h$p_value_adj<0.05) & (data_h$cell_count>10) & (data_h$est>0),]))
  overdispersed <- c(overdispersed,nrow(data_h[(data_h$z_score>0) & (data_h$p_value_adj<0.05) & (data_h$cell_count>10) & (data_h$est>0),]))
  
  # list of genes in human
  under_h_list <- data_h[(data_h$z_score<0) & (data_h$p_value_adj<0.05) & (data_h$cell_count>10) & (data_h$est>0),]$X
  over_h_list <- data_h[(data_h$z_score>0) & (data_h$p_value_adj<0.05) & (data_h$cell_count>10) & (data_h$est>0),]$X
  
  overlap_underdispersed <-c(overlap_underdispersed,length(intersect(under_c_list,under_h_list)))
}

species <- c("chimp","human","chimp","human","chimp","human")
replicate <- c("Rep1","Rep1","Rep2","Rep2","Rep3","Rep3")


cbind(species,replicate,underdispersed,overdispersed)
```

Overlap of significant gene across replicates.
```{r}
overlap_underdispersed <-c()
overlap_overdispersed <-c()
species <- c("human","chimp")
cell_types <-c("Cardiomyocytes","Undifferentiated1")
for (sp in species){
  for (cell in cell_types){
    output_name_1 = paste0(path, sp ,'_ase_cpm_', 'Rep1_',cell)
    output_name_2 = paste0(path, sp,'_ase_cpm_', 'Rep2_', cell)
    output_name_3 = paste0(path, sp,'_ase_cpm_', 'Rep3_', cell)
    data_1 <- read.csv(paste0(output_name_1,"_dispersion.csv"))
    data_2 <- read.csv(paste0(output_name_2,"_dispersion.csv"))
    data_3 <- read.csv(paste0(output_name_3,"_dispersion.csv"))
    
    # list of genes in Rep1
    under_1_list <- data_1[(data_1$z_score<0) & (data_1$p_value_adj<0.05) & (data_1$cell_count>10) & (data_1$est>0),]$X
    over_1_list <- data_1[(data_1$z_score>0) & (data_1$p_value_adj<0.05) & (data_1$cell_count>10) & (data_1$est>0),]$X
    
    # list of genes in Rep2
    under_2_list <- data_2[(data_2$z_score<0) & (data_2$p_value_adj<0.05) & (data_2$cell_count>10) & (data_2$est>0),]$X
    over_2_list <- data_2[(data_2$z_score>0) & (data_2$p_value_adj<0.05) & (data_2$cell_count>10) & (data_2$est>0),]$X
    
    # list of genes in Rep3
    under_3_list <- data_3[(data_3$z_score<0) & (data_3$p_value_adj<0.05) & (data_3$cell_count>10) & (data_3$est>0),]$X
    over_3_list <- data_3[(data_3$z_score>0) & (data_3$p_value_adj<0.05) & (data_3$cell_count>10) & (data_3$est>0),]$X
    
    overlap_underdispersed <-c(overlap_underdispersed, length(Reduce(intersect, list(under_1_list, under_2_list, under_3_list))))
                               
                          
    overlap_overdispersed <-c(overlap_overdispersed, length(Reduce(intersect, list(over_1_list, over_2_list, over_3_list))))
  }
}

species <- c("human","human","chimp","chimp")
cell_type <- c(cell_types,cell_types)

cbind(species,cell_type,overlap_underdispersed,overlap_overdispersed)
```


We count the overlap of significant genes within each replicate across species.
```{r}
overlap_underdispersed <-c()
overlap_overdispersed <-c()
cell_types <-c("Cardiomyocytes","Undifferentiated1")
for (cell in cell_types){
  for (rep in reps){
    output_name_c = paste0(path,'chimp_ase_cpm_', rep,'_',cell)
    output_name_h = paste0(path, 'human_ase_cpm_', rep, '_', cell)
    data_c <- read.csv(paste0(output_name_c,"_dispersion.csv"))
    data_h <- read.csv(paste0(output_name_h,"_dispersion.csv"))
    
    
    # list of genes in chimp
    under_c_list <- data_c[(data_c$z_score<0) & (data_c$p_value_adj<0.05) & (data_c$cell_count>10) & (data_c$est>0),]$X
    over_c_list <- data_c[(data_c$z_score>0) & (data_c$p_value_adj<0.05) & (data_c$cell_count>10) & (data_c$est>0),]$X
    
    
    # list of genes in human
    under_h_list <- data_h[(data_h$z_score<0) & (data_h$p_value_adj<0.05) & (data_h$cell_count>10) & (data_h$est>0),]$X
    over_h_list <- data_h[(data_h$z_score>0) & (data_h$p_value_adj<0.05) & (data_h$cell_count>10) & (data_h$est>0),]$X
    
    overlap_underdispersed <-c(overlap_underdispersed,length(intersect(under_c_list,under_h_list)))
    overlap_overdispersed <-c(overlap_overdispersed,length(intersect(over_c_list,over_h_list)))
  }
}

replicate <- c(reps,reps)
cell_type <- c("Cardiomyocytes","Cardiomyocytes","Cardiomyocytes","Undifferentiated 1","Undifferentiated 1","Undifferentiated 1")

cbind(cell_type,replicate,overlap_underdispersed,overlap_overdispersed)

```



