---
title: "Sharing of DV Genes in cis"
output: html_document
date: '2025-05-06'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# **Run mash**

We want to assess the sharing of DV genes in cis. We first ran memento for DV analysis to yield effect sizes and standard errors. We then used these effect sizes and standard errors as inputs for mash. Additionally, to compare a larger number of genes across conditions, we set the effect sizes to 0 and standard errors to $10^{16}$ for genes that were tested in one condition but not in another condition.

```{r, warning =FALSE, message=FALSE}
library(mashr)
library(ggplot2)
library(reshape2)
library(dplyr)
library(tidyr)
```


```{r, eval=FALSE, include=FALSE}
data_path = "/project2/gilad/awchen55/differentialDispersion/data/hybrid_data_MoM_results/hybrid_line_crossfilt_sharing/"

betahat <- read.csv(paste0(data_path, "betahat.csv"), row.names = 1)
sehat <- read.csv(paste0(data_path, "sehat.csv"), row.names = 1)
```




```{r, eval=FALSE, include=FALSE}
U.pca = cov_pca(data_mash,5,subset=strong)
print(names(U.pca))
U.ed = cov_ed(data_mash, U.pca, subset=strong)
```

```{r, eval=FALSE, include=FALSE}
m.ed = mash(data, U.ed)
print(get_loglik(m.ed),digits = 10)

```




```{r, warning=FALSE, message=FALSE, eval=FALSE, include=FALSE}
data_path = "/project2/gilad/awchen55/differentialDispersion/data/hybrid_data_MoM_results/hybrid_line_crossfilt_sharing/"

betahat <- read.csv(paste0(data_path, "betahat_ones.csv"), row.names = 1)
sehat <- read.csv(paste0(data_path, "sehat_inf.csv"), row.names = 1)

betahat_matrix <- as.matrix(betahat)
sehat_matrix   <- as.matrix(sehat)

data_mash <- mash_set_data(betahat_matrix, sehat_matrix)
m.1by1 = mash_1by1(data_mash)
strong = get_significant_results(m.1by1,0.05)


U.pca = cov_pca(data_mash,5,subset=strong)
U.ed = cov_ed(data_mash, U.pca, subset=strong)
# Combine covariance structures
Ulist = c(U.ed, U.pca)
m.ed_ones = mash(data_mash, Ulist = Ulist)

saveRDS(m.ed_ones, file = paste0(data_path ,"hybrid_crossfilt_mash_results_ones.rds"))
```

```{r, warning=FALSE, message=FALSE, eval=FALSE, include=FALSE}
data_path = "/project2/gilad/awchen55/differentialDispersion/data/hybrid_data_MoM_results/hybrid_line_crossfilt_sharing/"

betahat <- read.csv(paste0(data_path, "betahat_no_nan.csv"), row.names = 1)
sehat <- read.csv(paste0(data_path, "sehat_no_nan.csv"), row.names = 1)

betahat_matrix <- as.matrix(betahat)
sehat_matrix   <- as.matrix(sehat)

data_mash <- mash_set_data(betahat_matrix, sehat_matrix)
m.1by1 = mash_1by1(data_mash)
strong = get_significant_results(m.1by1,0.05)


U.pca = cov_pca(data_mash,5,subset=strong)
U.ed = cov_ed(data_mash, U.pca, subset=strong)
# Combine covariance structures
Ulist = c(U.ed, U.pca)
m.ed_no_nan = mash(data_mash, Ulist = Ulist)

saveRDS(m.ed_no_nan, file = paste0(data_path ,"hybrid_crossfilt_mash_results_no_nan.rds"))
```

```{r, eval=FALSE, include=FALSE}
length(get_significant_results(m.ed_no_nan,conditions=1))
sig_genes <- get_significant_results(m.ed_no_nan,conditions=1)
lfsr_thresh=0.05
i=1
j=2
factor = 0.5
sig_i=get_significant_results(m.ed_no_nan,thresh=lfsr_thresh,conditions = i)
sig_j=get_significant_results(m.ed_no_nan,thresh=lfsr_thresh,conditions = j)
a=union(sig_i,sig_j)
ratio = get_pm(m.ed_no_nan)[a,i]/get_pm(m.ed_no_nan)[a,j]##divide effect sizes

length(ratio[(ratio>factor & ratio<(1/factor))==TRUE])
#S[i,j]=mean(ratio>factor & ratio<(1/factor))
# for(i in length(get_significant_results(m.ed_no_nan,conditions=1)){
#   print(i)
# }
```



```{r, eval=FALSE, include=FALSE}
n_conditions = 15
shared_genes_list <- list()
lfsr_thresh=0.05
factor = 0.5

all_shared_genes <-c()
for (i in 1:(n_conditions - 1)) {
  for (j in (i + 1):n_conditions) {
    sig_i=get_significant_results(m.ed_no_nan,thresh=lfsr_thresh,conditions = i)
    sig_j=get_significant_results(m.ed_no_nan,thresh=lfsr_thresh,conditions = j)
    a=union(sig_i,sig_j)
    ratio = get_pm(m.ed_no_nan)[a,i]/get_pm(m.ed_no_nan)[a,j]##divide effect sizes
    
    overlap_genes <- names(ratio[(ratio>factor & ratio<(1/factor))==TRUE])
    
    pair_name <- paste0('celltype', i-1,"_vs_",'celltype',j-1)
    shared_genes_list[[pair_name]] <- overlap_genes
    
    all_shared_genes <- c(all_shared_genes, overlap_genes)
  }
}

```

```{r, eval=FALSE, include=FALSE}
unique_shared_genes <- unique(all_shared_genes)
length(get_significant_results(m.ed_no_nan,thresh=lfsr_thresh))
length(unique_shared_genes)
shared_genes_all <- c(names(get_significant_results(m.ed_no_nan,thresh=lfsr_thresh)),unique_shared_genes)
length(unique(shared_genes_all))
```



```{r, eval=FALSE, include=FALSE}
# pvalue test
pvalues_ones <- read.csv(paste0(data_path, "pvalues_ones.csv"), row.names = 1)
betahat_zeros <- read.csv(paste0(data_path, "betahat_zeros.csv"), row.names = 1)
#zSE = abs(betahat_zeros/ qnorm(pvalues_ones/2))
zSE <- c()
for(i in 1:15){
  zSEi <- abs(betahat_zeros[[i]]/ qnorm(pvalues_ones[[i]]/2))
  zSE <- cbind(zSE,zSEi)
}
zSE[is.na(zSE)] <- 1e16
rownames(zSE) <- rownames(betahat_zeros)
colnames(zSE) <- colnames(betahat_zeros)
#zSE
zSE[7527,2] = sehat[7527,2]
betahat_matrix <- as.matrix(betahat_zeros)
sehat_matrix   <- as.matrix(zSE)

data_mash <- mash_set_data(betahat_matrix, sehat_matrix)
m.1by1 = mash_1by1(data_mash)
strong = get_significant_results(m.1by1,0.05)


U.pca = cov_pca(data_mash,5,subset=strong)
U.ed = cov_ed(data_mash, U.pca, subset=strong)
# Combine covariance structures
Ulist = c(U.ed, U.pca)
m.ed_zSE = mash(data_mash, Ulist = Ulist)

saveRDS(m.ed_zSE, file = paste0(data_path ,"hybrid_crossfilt_mash_results_zeros_zSE.rds"))
```


```{r, eval=FALSE}
data_path = "/project2/gilad/awchen55/differentialDispersion/data/hybrid_data_MoM_results/hybrid_line_crossfilt_sharing/"

betahat <- read.csv(paste0(data_path, "betahat_zeros.csv"), row.names = 1)
sehat <- read.csv(paste0(data_path, "sehat_inf.csv"), row.names = 1)

betahat_matrix <- as.matrix(betahat)
sehat_matrix   <- as.matrix(sehat)

data_mash <- mash_set_data(betahat_matrix, sehat_matrix)
m.1by1 = mash_1by1(data_mash)
strong = get_significant_results(m.1by1,0.05)


U.pca = cov_pca(data_mash,5,subset=strong)
U.ed = cov_ed(data_mash, U.pca, subset=strong)
# Combine covariance structures
Ulist = c(U.ed, U.pca)
m.ed_zeros = mash(data_mash, Ulist = Ulist)

#saveRDS(m.ed_zeros, file = paste0(data_path ,"hybrid_crossfilt_mash_results_zeros.rds"))

```


```{r, eval=FALSE, include=FALSE}
# Beta with ones
data_path = "/project2/gilad/awchen55/differentialDispersion/data/hybrid_data_MoM_results/hybrid_line_crossfilt_sharing/"

betahat <- read.csv(paste0(data_path, "betahat_ones.csv"), row.names = 1)
sehat <- read.csv(paste0(data_path, "sehat_inf.csv"), row.names = 1)

betahat_matrix <- as.matrix(betahat)
sehat_matrix   <- as.matrix(sehat)

data_mash <- mash_set_data(betahat_matrix, sehat_matrix)
m.1by1 = mash_1by1(data_mash)
strong = get_significant_results(m.1by1,0.05)


U.pca = cov_pca(data_mash,5,subset=strong)
U.ed = cov_ed(data_mash, U.pca, subset=strong)
# Combine covariance structures
Ulist = c(U.ed, U.pca)
m.ed_ones = mash(data_mash, Ulist = Ulist)

saveRDS(m.ed_ones, file = paste0(data_path ,"hybrid_crossfilt_mash_results_ones.rds"))

```



```{r, eval=FALSE, include=FALSE}
n_conditions = 15
shared_genes_list <- list()
lfsr_thresh=0.05
factor = 0.5

all_shared_genes <-c()
for (i in 1:(n_conditions - 1)) {
  for (j in (i + 1):n_conditions) {
    # get significant genes for each condition
    sig_i=get_significant_results(m.ed_ones,thresh=lfsr_thresh,conditions = i)
    sig_j=get_significant_results(m.ed_ones,thresh=lfsr_thresh,conditions = j)
    a=union(sig_i,sig_j)
    
    # get ratio of effect sizes
    ratio = get_pm(m.ed_ones)[a,i]/get_pm(m.ed_ones)[a,j]##divide effect sizes
    
    # determine if ratio falls within factor
    overlap_genes <- names(ratio[(ratio>factor & ratio<(1/factor))==TRUE])
    
    # create list of shared genes for each pairwise combo
    pair_name <- paste0('celltype', i-1,"_vs_",'celltype',j-1)
    shared_genes_list[[pair_name]] <- overlap_genes
    
    # list of all shared genes
    all_shared_genes <- c(all_shared_genes, overlap_genes)
  }
}

```

```{r, eval=FALSE, include=FALSE}
unique_shared_genes <- unique(all_shared_genes)
length(get_significant_results(m.ed_ones,thresh=lfsr_thresh))
length(unique_shared_genes)
shared_genes_all <- c(names(get_significant_results(m.ed_ones,thresh=lfsr_thresh)),unique_shared_genes)
length(unique(shared_genes_all))
```

# **Plots**
## ***Total DV Genes***
We first assess the number of significant DV genes within each cell type. We note DV genes that exhibit lower variability with respect to each species. 
```{r}
data_path = "/project2/gilad/awchen55/differentialDispersion/data/hybrid_data_MoM_results/hybrid_line_crossfilt_sharing/"
lfsr_thresh=0.05
betahat <- read.csv(paste0(data_path, "betahat_zeros.csv"), row.names = 1)
sehat <- read.csv(paste0(data_path, "sehat_inf.csv"), row.names = 1)
m.ed_zeros <- readRDS(paste0(data_path ,"hybrid_crossfilt_mash_results_zeros.rds"))

genes_higher_variability_chimps <- list()
genes_higher_variability_humans <- list()
for (cell_type in colnames(betahat)) {
  genes_higher_variability_humans[[cell_type]] <- rownames(betahat[betahat[[cell_type]] > 0, ])
  genes_higher_variability_chimps[[cell_type]] <- rownames(betahat[betahat[[cell_type]] < 0, ])
}
num_genesc <- c()
num_genesh <- c()
dv_total <-c()
for (i in 1:15){
  sig_i=list(names(get_significant_results(m.ed_zeros,thresh=lfsr_thresh,conditions = i)))
  overlap_chimp <- intersect(sig_i[[1]], genes_higher_variability_chimps[[i]])
  overlap_human <- intersect(sig_i[[1]], genes_higher_variability_humans[[i]])
  num_genesc <- c(num_genesc, length(overlap_chimp))
  num_genesh <- c(num_genesh, length(overlap_human))
  dv_total <-c(dv_total,length(sig_i[[1]]))
}

# num_genesc
# num_genesh
# dv_total
```

```{r}
dv_gene_totals <- cbind(num_genesh,num_genesc)
colnames(dv_gene_totals) <- c('human','chimpanzee')
rownames(dv_gene_totals) <- colnames(betahat)
#dv_gene_totals
data_df <- melt(dv_gene_totals, varnames = c("Cell_Type", "Species"), value.name = "DV_Genes")

# Plot the bar graph with data labels
p1 <- ggplot(data_df, aes(x = Cell_Type, y = DV_Genes, fill = Species)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8), width = 0.6) +
  geom_text(aes(label = DV_Genes), 
            position = position_dodge(width = 0.8), 
            vjust = -0.5, 
            size = 2) +
  labs(title = "DV Genes with Lower Variability",
       x = "",
       y = "# DV Genes (LFSR < 0.05)") +
  theme_minimal() +
  theme(
         plot.title = element_text(size = 10, face = 'bold', hjust = 0.5),
      panel.grid = element_blank(),
      axis.line = element_line(color = "black", size = 0.4),
      panel.border = element_rect(color = "black", fill = NA, size = 0.8),
      axis.title.x = element_text(size = 8),
      axis.title.y = element_text(size = 8),
      axis.text.x = element_text(angle = 90, hjust = 1, size=8),
      axis.text.y = element_text(size = 8),
      legend.position = c(0.98, 0.95),    # Position in the top-right corner inside the plot
      legend.justification = c("right", "top"),
      legend.title = element_blank(),    # Remove the legend title
      legend.background = element_rect(fill = "white", color = NA, size = 0.2),
      legend.text = element_text(size = 7),
      legend.key.size = unit(0.3, "cm")
      )
p1
```

## ***Shared vs Unique Genes***
Next we assess pairwise sharing of DV genes across cell types. For every cell type combination we check sharing by seeing if DV genes share the same sign and if the ratio of their effect sizes across two conditions are within a factor of 0.5.
```{r}
n_conditions = 15
shared_genes_list <- list()
lfsr_thresh=0.05
factor = 0.5

all_shared_genes <-c()
for (i in 1:(n_conditions - 1)) {
  for (j in (i + 1):n_conditions) {
    sig_i=get_significant_results(m.ed_zeros,thresh=lfsr_thresh,conditions = i)
    sig_j=get_significant_results(m.ed_zeros,thresh=lfsr_thresh,conditions = j)
    a=union(sig_i,sig_j)
    ratio = get_pm(m.ed_zeros)[a,i]/get_pm(m.ed_zeros)[a,j]##divide effect sizes
    
    overlap_genes <- names(ratio[(ratio>factor & ratio<(1/factor))==TRUE])
    
    pair_name <- paste0('celltype', i-1,"_vs_",'celltype',j-1)
    shared_genes_list[[pair_name]] <- overlap_genes
    
    all_shared_genes <- c(all_shared_genes, overlap_genes)
  }
}

```

```{r}
unique_shared_genes <- unique(all_shared_genes)
length(get_significant_results(m.ed_zeros,thresh=lfsr_thresh))
length(unique_shared_genes)
shared_genes_all <- c(names(get_significant_results(m.ed_zeros,thresh=lfsr_thresh)),unique_shared_genes)
length(unique(shared_genes_all))
```
```{r}
data_df <- data.frame(
  Condition = c("Unique", "Shared"),
  Value = c(length(unique(shared_genes_all)) - length(unique_shared_genes), length(unique_shared_genes))
)

# Plot the bar graph
p2 <- ggplot(data_df, aes(x = Condition, y = Value, fill = Condition)) +
  geom_bar(stat = "identity", width = 0.6) +
  geom_text(aes(label = Value), vjust = -0.5, size = 3) +
  scale_fill_manual(values = c("Shared" = "forestgreen", "Unique" = "royalblue")) +
  labs(
    title = "Shared vs Unique DV Genes",
    x = "",
    y = "# DV Genes"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 10, face = 'bold', hjust = 0.5),
    axis.text.x = element_text(size = 8),
    axis.text.y = element_text(size = 8),
    axis.title.y = element_text(size = 8),
    panel.grid = element_blank(),
      axis.line = element_line(color = "black", size = 0.4),
      panel.border = element_rect(color = "black", fill = NA, size = 0.8),
    legend.position = "none"  # Hide the legend since labels are clear
  )
p2
```

## ***Types of DV Genes***

### *Run mash on DE*
We first run mash on the effect sizes and standard errors of DE analysis using memento. 
```{r}
data_path = "/project2/gilad/awchen55/differentialDispersion/data/hybrid_data_MoM_results/hybrid_line_crossfilt_sharing/"

betahat <- read.csv(paste0(data_path, "de_betahat_zeros.csv"), row.names = 1)
sehat <- read.csv(paste0(data_path, "de_sehat_inf.csv"), row.names = 1)

betahat_matrix <- as.matrix(betahat)
sehat_matrix   <- as.matrix(sehat)

data_mash <- mash_set_data(betahat_matrix, sehat_matrix)
m.1by1 = mash_1by1(data_mash)
strong = get_significant_results(m.1by1,0.05)


U.pca = cov_pca(data_mash,5,subset=strong)
U.ed = cov_ed(data_mash, U.pca, subset=strong)
# Combine covariance structures
Ulist = c(U.ed, U.pca)
m.ed_zeros_de = mash(data_mash, Ulist = Ulist)

#saveRDS(m.ed_zeros_de, file = paste0(data_path ,"hybrid_crossfilt_mash_results_zeros_de.rds"))
```

### *Categorize Types of DV genes*
We categorize DV genes into three types: DV genes that are only DV, DV genes that are also DE with the same sign, or DV/DE genes with opposite signs.
```{r}

betahat_dv <- read.csv(paste0(data_path, "betahat_zeros.csv"), row.names = 1)
betahat_de <- read.csv(paste0(data_path, "de_betahat_zeros.csv"), row.names = 1)

genes_lower_variability_chimps <- list()
genes_lower_variability_humans <- list()
cell_types <- colnames(betahat)
dv_only_h <- c()
dv_de_same_h <- c()
dv_de_opp_h <- c()
h_values <- c()
dv_only_c <- c()
dv_de_same_c <- c()
dv_de_opp_c <- c()
c_values <- c()
for (i in 1:length(cell_types) ){
  cell_type = cell_types[i]
  # get signs of dv
  genes_lower_variability_h_dv <- rownames(betahat_dv[betahat_dv[[cell_type]] > 0, ])
  genes_lower_variability_c_dv <- rownames(betahat_dv[betahat_dv[[cell_type]] < 0, ])
  
  # get signs of de
  genes_lower_expression_h <- rownames(betahat_de[betahat_de[[cell_type]] > 0, ])
  genes_lower_expression_c <- rownames(betahat_de[betahat_de[[cell_type]] < 0, ])
  
  # sig dv and de
  sig_de <- names(get_significant_results(m.ed_zeros_de,thresh=0.05, conditions = i))
  sig_dv <- names(get_significant_results(m.ed_zeros,thresh=0.05, conditions = i))
  
  # get signs of sig dv and de
  dv_low_h <- intersect(sig_dv,genes_lower_variability_h_dv)
  dv_low_c <- intersect(sig_dv,genes_lower_variability_c_dv)
  de_low_h <- intersect(sig_de,genes_lower_expression_h)
  de_low_c <- intersect(sig_de,genes_lower_expression_c)
  
  # categorize dv/de genes
  dv_de_same_sign_h <- intersect(dv_low_h,de_low_h)
  dv_de_opp_sign_h <- intersect(dv_low_h,de_low_c)
  
  dv_de_same_sign_c <- intersect(dv_low_c,de_low_c)
  dv_de_opp_sign_c <- intersect(dv_low_c,de_low_h)
  
  
  # count dv genes for humans low var
  dv_de_same_h <- c(dv_de_same_h,length(dv_de_same_sign_h))
  dv_de_opp_h <- c(dv_de_opp_h,length(dv_de_opp_sign_h))
  dv_only_h <- c(dv_only_h,length(dv_low_h) - length(dv_de_same_sign_h) - length(dv_de_opp_sign_h))
  h_values <- c(h_values,length(dv_de_same_sign_h),length(dv_de_opp_sign_h),length(dv_low_h) - length(dv_de_same_sign_h) - length(dv_de_opp_sign_h))
  # count dv genes for chimps low var
  dv_de_same_c <- c(dv_de_same_c,length(dv_de_same_sign_c))
  dv_de_opp_c <- c(dv_de_opp_c,length(dv_de_opp_sign_c))
  dv_only_c <- c(dv_only_c,length(dv_low_c) - length(dv_de_same_sign_c) - length(dv_de_opp_sign_c))
  c_values <- c(c_values,length(dv_de_same_sign_c),length(dv_de_opp_sign_c),length(dv_low_c) - length(dv_de_same_sign_c) - length(dv_de_opp_sign_c))
}

```

```{r}

# Data setup
cell_types <- paste0('celltype', 0:14)

# Combine into a data frame
data <- data.frame(
  CellType = rep(cell_types, each = 3),
  Category = rep(c('DE/DV Same Signs', 'DE/DV Opposite Signs','DV Only'), times = 15),
  Value = h_values
)

# Convert to percentage within each cell type
data <- data %>%
  group_by(CellType) %>%
  mutate(Percentage = Value / sum(Value) * 100)

# Ensure the order of cell types is preserved numerically
data$CellType <- factor(data$CellType, levels = paste0('celltype', 0:14))

# Plot
p3 <- ggplot(data, aes(x = CellType, y = Percentage, fill = Category)) +
  geom_bar(stat = 'identity') +
  scale_fill_manual(values = c('blue3', 'orange2', 'purple2')) +
  labs(
    title = 'Low Variability in Humans',
    y = '% of DV Genes',
    x = ''
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 10, face = 'bold', hjust = 0.5),
    axis.text.x = element_text(size = 8, angle = 90, hjust = 1),
    axis.text.y = element_text(size = 8),
    axis.title.y = element_text(size = 8),
    panel.grid = element_blank(),
    #legend.position = "none"  # Hide the legend since labels are clear
  )
p3
```

```{r}
# Data setup
cell_types <- paste0('celltype', 0:14)

# Combine into a data frame
data <- data.frame(
  CellType = rep(cell_types, each = 3),
  Category = rep(c('DE/DV Same Signs', 'DE/DV Opposite Signs','DV Only'), times = 15),
  Value = c_values
)

# Convert to percentage within each cell type
data <- data %>%
  group_by(CellType) %>%
  mutate(Percentage = Value / sum(Value) * 100)

# Ensure the order of cell types is preserved numerically
data$CellType <- factor(data$CellType, levels = paste0('celltype', 0:14))

# Plot
p4 <- ggplot(data, aes(x = CellType, y = Percentage, fill = Category)) +
  geom_bar(stat = 'identity') +
  scale_fill_manual(values = c('blue3', 'orange2', 'purple2')) +
  labs(
    title = 'Low Variability in Chimpanzees',
    y = '% of DV Genes',
    x = ''
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 10, face = 'bold', hjust = 0.5),
    axis.text.x = element_text(size = 8, angle = 90, hjust = 1),
    axis.text.y = element_text(size = 8),
    axis.title.y = element_text(size = 8),
    panel.grid = element_blank(),
    #legend.position = "none"  # Hide the legend since labels are clear
  )
p4
```






